SCPCL000822 - Validation of tumor cells

### 2024-06-05 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>

# copykat output
copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
  # only select no_ref classification and rename to be consistent with naming of other classifications
  dplyr::select(barcodes, copykat_classification = no_ref) |> 
  dplyr::mutate(
    copykat_classification = dplyr::case_when(
      copykat_classification == "aneuploid" ~ "Tumor",
      copykat_classification == "diploid" ~ "Normal",
      TRUE ~ copykat_classification
    )
  )

@@

# pull out the UMAP coordinates and make a data frame to use for plotting
coldata_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
  # add in marker gene classifications
  dplyr::left_join(marker_gene_results_df) |> 
  # add in copykat results 
  dplyr::left_join(copykat_results_df) |> 
  # add in infercnv results and metadata
  dplyr::left_join(infercnv_results_df) |> 
  dplyr::left_join(infercnv_metadata_df) |> 
  # gene set scores 
  dplyr::left_join(geneset_scores_df)

@@

```{r}
# look at which cells are tumor/normal
# be sure to remove any duplicates 
coldata_df |> 
  dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
  unique() |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

@@

```{r}
# plot heatmap of marker genes 
ComplexHeatmap::Heatmap(
  marker_gene_heatmap,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "Marker gene symbol",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  show_column_names = FALSE,
  bottom_annotation = annotation
)
```
As expected, the tumor cells appear to have more marker genes that are expressed than normal cells.

@@

```{r, fig.height=10}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bounds = c(0,1), bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

>>>>

# copykat output
copykat_results_df <- readr::read_tsv(copykat_predictions_file) |> 
  # only select no_ref classification and rename to be consistent with naming of other classifications
  dplyr::select(barcodes, copykat_classification = no_ref) |> 
  dplyr::mutate(
    copykat_classification = dplyr::case_when(
      copykat_classification == "aneuploid" ~ "Tumor",
      copykat_classification == "diploid" ~ "Normal",
      .default = copykat_classification
    )
  )

@@

# pull out the UMAP coordinates and make a data frame to use for plotting
coldata_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = "ensembl_gene_id") |>
  # get rid of excess columns
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression) |> 
  # add in marker gene classifications
  dplyr::left_join(marker_gene_results_df, by = "barcodes") |> 
  # add in copykat results 
  dplyr::left_join(copykat_results_df, by = "barcodes") |> 
  # add in infercnv results and metadata
  dplyr::left_join(infercnv_results_df, by = "barcodes") |> 
  dplyr::left_join(infercnv_metadata_df, by = "barcodes") |> 
  # gene set scores 
  dplyr::left_join(geneset_scores_df, by = "barcodes")

@@

```{r}
# look at which cells are tumor/normal
# be sure to remove any duplicates 
coldata_df |> 
  dplyr::select(UMAP1, UMAP2, barcodes, combined_cnv_classification) |>
  dplyr::distinct() |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

@@

```{r}
# plot heatmap of marker genes 
ComplexHeatmap::Heatmap(
  marker_gene_heatmap,
  col = circlize::colorRamp2(c(0, 1), colors = c("white", "#00274C")),
  border = TRUE,
  ## Row parameters
  cluster_rows = TRUE,
  row_title = "Marker gene symbol",
  row_title_gp = grid::gpar(fontsize = 12),
  row_title_side = "left",
  row_names_side = "left",
  row_dend_side = "right",
  row_names_gp = grid::gpar(fontsize = 10),
  ## Column parameters
  cluster_columns = TRUE,
  show_column_names = FALSE,
  bottom_annotation = annotation
)
```

As expected, the tumor cells appear to have more marker genes that are expressed than normal cells.

@@

```{r, fig.height=10}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```
